---
title: "Answer Key for Examples"
author: "Kevin Rozmiarek"
date: "9/1/2020"
output: html_document
---

# Questions

1. Let me add that the approximate temperature dependence of the fractionation factor for water evaporation in reality follows this expression $\alpha = 0.9822e^{(7.9448 /T)}$ Show the curve for $\delta$ value of the liquid as a function of remaining unevaporated liquid. The starting isotope value of the water is $\delta = -20$ and the water is held at 35°C. Why do you think these numbers are so high?

Answer: 

```{r message=FALSE}
library(tidyverse)
plot_Data_Neg_twenty <- tibble(
  remaining_Water_Fraction = rev(seq(0.01,1,0.01)),
  temperature = -35
)

plot_Data_Neg_twenty <- plot_Data_Neg_twenty %>%
  mutate(d18O = (-20 + 1000)*(remaining_Water_Fraction^(0.9822*exp(7.9448/temperature)-1))-1000)
```

```{r}
library(ggplot2)
theplot <- ggplot(plot_Data_Neg_twenty, aes(x = remaining_Water_Fraction,y = d18O)) + 
  geom_point(aes(x = remaining_Water_Fraction,y = d18O))

plot(theplot)
```

The number is high because reality includes condensation and an experiment that tests for fractionation factor would observe that.

2. Speculate on how you would determine the isotope value for two independent fractionations happening at once.

Answer: You could add the factors linearily such as:

$$
\delta_{l} = (\delta_{l,0} + 1000)(f^{\alpha_{1}-1} + f^{\alpha_{2}-1}) - 1000
$$