---
title: "Isotopes Fractionation"
subtitle: "Isoptopes are isoDOPE"
author: "Kevin Rozmiarek"
date: "Last knitted `r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document: 
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: true
    toc_depth: 3
    code_folding: show
editor_options:
  chunk_output_type: inline
---

# Learning Goals

1. Understand the role of **fractionation factors** in seperating isotopes in different reservoirs

2. Understand the difference between **equilibrium isotope fractionation** and **kinetic isotopic fractionation**

3. Be able to numerically demostrate the seperation of two isotopic reservoirs by evaporation. 

# Background

Because we just finished our lesson on what isotopes are and their notation *cough*, it is time to talk about **isotope fractionation**. **Isotope fractionation** is the seperation or partioning of a reservoir(s) stable isotopes due to a processe that use masses differently. Some processes depend on the relative differences in the masses of different isotopes (**equilibrium isotopic fractionation**). Other processes depend on rate-dependent transfers or site-specific chemical binding effects (**kinetic isotopic fractionation**).

Today we are going to talk about the evaporation of water which happens to be an example of a rate-dependent kinetic fractionation. Well... its kinetic in very dry air. Let's not go there for now. The fractionation from a process is described by using an often empirically derived non-dimensional parameter called a **fractionation factor**. Evaporation is refer to a type of fractionation called a **Rayleigh distillation**. In an **open system**, isotopes will not return to the original reservoir once evaporated. At this point, we can write out a general form for a Rayleigh distillation that will tell us the isotopic ratio of the reservoir as a function of the remaining amount that hasn't been evaporated.

$$
\frac{R}{R_{0}} = f^{\alpha-1}
$$

where *R* = ratio of the isotopes (e.g., 18O/16O) left in the reservoir, *Ro* = initial ratio, *f* = fraction of material remaining in the reservoir, and *α* is the fractionation factor.

We can do some algebra now and rewrite this in $\delta$ notation. I will spare you the trouble.

$$
\delta_{l} = (\delta_{l,0} + 1000)f^{\alpha-1} - 1000
$$

where *o* stands for the initial δ-value of the liquid (*l*) for the case when α is the fractionation factor between vapor and liquid αvapor-liquid, and f is the fraction of liquid remaining.

# Example

The emprically derived fractionation factor associated with the phase transition during evaporation, vapor-liquid = 0.990 at 25°C. Think of this as saying that the new vapor is 10‰ lighter than the residual water. Now let's use the above equation to calculate the curve for a starting water delta value of $\delta = -10$.

```{r, message=FALSE}
library(tidyverse)

# session info
data <- tibble(
  `d18O_0` = seq(-10,30,10),
  alpha = 0.990,
)

#Adding column for calculated term
data <- data %>%
  mutate(d18O_50 = (d18O_0 + 1000)*(0.5^(alpha-1))-1000)

#Data for plotting at one start value of d180
plot_Data_Neg_ten <- tibble(
  remaining_Water_Fraction = rev(seq(0.01,1,0.01)),
  alpha = 0.990
)

#Adding column for calculated term
plot_Data_Neg_ten <- plot_Data_Neg_ten %>%
  mutate(d18O = (-10 + 1000)*(remaining_Water_Fraction^(alpha-1))-1000)
```

```{r}
library(ggplot2)
theplot <- ggplot(plot_Data_Neg_ten, aes(x = remaining_Water_Fraction,y = d18O)) + 
  geom_point(aes(x = remaining_Water_Fraction,y = d18O))

plot(theplot)
```

# Questions

1. Let me add that the approximate temperature dependence of the fractionation factor for water evaporation follows this expression $\alpha = 0.9822e^{(7.9448 /T)}$ Show the curve for $\delta$ value of the liquid as a function of remaining unevaporated liquid. The starting isotope value of the water is $\delta = -20$ and the water is held at 35°C. Why do you think these numbers are so high?

2. Speculate on how you would determine the isotope value for two independent fractionations happening at once.